---
title: "ICTV validate proposals"
params:
  # inputs
  prev_taxa_fname: ./current_msl/taxonomy_node_names.all.txt
  rank_fname: ./current_msl/taxonomy_level.txt
  cv_xlsx: ./TP_Template_Excel_module_2022_v2.xlsx
  cv_sheet: "Menu Items (Do not change)"
  zip_dir: ./zips
  # set to false when editing .xlsx files to fix
  unzipTop:    FALSE
  unzipPro:    FALSE
  # output files
  tmp_dir: ./tmp
  pro_dir: ./pro_zips
  dest_msl: 38
  merged:  load_next_msl.txt
  status:  merged_status.txt
  
output: html_document
---

This script 
  * load previous MSL data
  * scans params$zip_dir for *.zip (proposal zips)
  * iterates over each proposal zip
    * unzips it into params$tmp_dir
    * reads the proposal.xlsx file into a data frame
    * merge each row (change)
      * match up to existing taxon
      * QC data, etc.
      * implement change in new MSL data frame
      * record errors, warnings, etc
      * create merged change set in one merged dataframe
  * write merged proposal data frame to a Unicode(UTF-16LE) TSV file (params$merged) that can be loaded into MSSQL on Windows using "Import Data...." 
  * write a status sheet listing parsing and QC success/fail status for each proposal (params$status)
  * write new MSL load & updates to prev_msl.out_*
  
```{r setup, include=FALSE}
library(yaml)
library(tidyverse)
library(readxl)
#library(gtools) # for mixedsort/mixedorder

library(knitr) #kable
# debug - echo everything
knitr::opts_chunk$set(echo = TRUE)

```

# load previous MSLs
```{r load prev MSL}

taxonomy_node_names = c("taxnode_id","parent_id","tree_id","msl_release_num","level_id","name","ictv_id","molecule_id","abbrev_csv","genbank_accession_csv","genbank_refseq_accession_csv","refseq_accession_csv","isolate_csv","notes","is_ref","is_official","is_hidden","is_deleted","is_deleted_next_year","is_typo","is_renamed_next_year","is_obsolete","in_change","in_target","in_filename","in_notes","out_change","out_target","out_filename","out_notes","start_num_sort","row_num","filename","xref","realm_id","realm_kid_ct","realm_desc_ct","subrealm_id","subrealm_kid_ct","subrealm_desc_ct","kingdom_id","kingdom_kid_ct","kingdom_desc_ct","subkingdom_id","subkingdom_kid_ct","subkingdom_desc_ct","phylum_id","phylum_kid_ct","phylum_desc_ct","subphylum_id","subphylum_kid_ct","subphylum_desc_ct","class_id","class_kid_ct","class_desc_ct","subclass_id","subclass_kid_ct","subclass_desc_ct","order_id","order_kid_ct","order_desc_ct","suborder_id","suborder_kid_ct","suborder_desc_ct","family_id","family_kid_ct","family_desc_ct","subfamily_id","subfamily_kid_ct","subfamily_desc_ct","genus_id","genus_kid_ct","genus_desc_ct","subgenus_id","subgenus_kid_ct","subgenus_desc_ct","species_id","species_kid_ct","species_desc_ct","taxa_kid_cts","taxa_desc_cts","inher_molecule_id","left_idx","right_idx","node_depth","lineage","cleaned_name","cleaned_problem","flags","_numKids","_out_target_parent","_out_target_name","rank","tree","realm","subrealm","kingdom","subkingdom","phylum","subphylum","class","subclass","order","suborder","family","subfamily","genus","subgenus","species","molecule","inher_molecule")
prevMSLs=read.delim(file=params$prev_taxa_fname,header=FALSE,stringsAsFactors=TRUE,col.names=taxonomy_node_names)

# CV loads
rankCV = read.delim(file=params$rank_fname,header=TRUE, stringsAsFactors=TRUE,na.strings=c("NULL"))
rownames(rankCV) = rankCV$id

```
# Load CVs for Proposal QC
```{r load cvs}
cvDf = data.frame(read_excel(params$cv_xlsx,sheet = params$cv_sheet,col_names = FALSE))
#cvDf = data.frame(trib[,])  # remove "select one" line

cvList=list()
for(cv_col in 1:ncol(cvDf)) {
  cv_name = cvDf[1,cv_col]
  cv = cvDf[,cv_col][-1]
  cvList[[cv_name]]=c(cv[!is.na(cv)],NA)
}

# map to output names
cvNameMap = c(
  "Genome coverage"=    "isComplete",
  "Genome composition"= "molecule",
  "Host/Source"=        "hostSource",
  "Change"=             "change",
  "Rank"=               "rank"  
)

names(cvList)=cvNameMap[names(cvList)]
```

# scan and extract zip files
```{r scan zip files}
total_proposals=0
topZipDf = data.frame(zip=list.files(path=params$zip_dir, full.names=TRUE, pattern="*.zip") )
rownames(topZipDf) = topZipDf$zip
topZipDf$basename = gsub(".*/([^/]+).zip$","\\1",topZipDf$zip)
topZipDf$proFolder = file.path(params$pro_dir,topZipDf$basename)

zipDf = setNames(data.frame(matrix(ncol = 5, nrow = 0)), c("topZip","proZip","proZipPath","proBase","xlsx"))

for(zip in topZipDf$zip) {
  # zip=topZipDf$zip[2] # debug
  cat("Scanning ", zip, "\n")
  # get toc
  contents = unzip(zipfile=zip, list=TRUE)$Name 
  # filter out MACOSX system files
  contents = grep(contents, pattern="__MACOSX", value = T, invert = T)
  # filter out docx-only zips (2021.015P.A.v1.Viroid_demarcation_criteria.docx.zip)
  contents = grep(contents, pattern=".*.docx.zip", value = T, invert = T)
  # extract contents into TMP
  if( params$unzipTop) { unzip(zipfile=zip, exdir=topZipDf[zip,"proFolder"], files = contents) }
  cat("\t",paste0(contents, collapse = "\n\t"), "\n")
  
  # build frame of file names...
  tocDf = data.frame(topZip=zip, 
                     proZip=contents, 
                     proPath=file.path(topZipDf[zip,"proFolder"],contents), 
                     proBase=gsub(".zip$","",contents), 
                     xlsxDir=file.path(params$tmp_dir,topZipDf[zip,"basename"],gsub(".zip$","",contents)),
                     xlsx=gsub(".zip$",".xlsx",contents)
                     )
  # append to total list
  zipDf = rbind(zipDf, tocDf)
  
}
rownames(zipDf) = zipDf$proZip
cat("# TOTAL PROPOSALS = ", nrow(zipDf), "\n\n")
```

## QC setup
```{r qc_setup}
#
# expected input header rows 2021
#

# dput(unname(as.vector(df[1,])))
xlsx_row2=c("CURRENT TAXONOMY", NA_character_, NA_character_, 
    NA_character_, NA_character_, NA_character_, NA_character_, 
    NA_character_, NA_character_, NA_character_, NA_character_, 
    NA_character_, NA_character_, NA_character_, NA_character_, 
    NA_character_, "PROPOSED TAXONOMY", NA_character_, NA_character_, 
    NA_character_, NA_character_, NA_character_, NA_character_, 
    NA_character_, NA_character_, NA_character_, NA_character_, 
    NA_character_, NA_character_, NA_character_, NA_character_, 
    NA_character_, NA_character_, NA_character_, NA_character_, 
    NA_character_, NA_character_, NA_character_, "SPECIFY PROPOSED CHANGE", 
    NA_character_, "COMMENTS", NA, NA, NA, NA_character_, NA_character_, 
    NA_character_, NA_character_, NA_character_)

# dput(unname(as.vector(df[2,])))
xlsx_row3=c("Realm", "Subrealm", "Kingdom", "Subkingdom", 
    "Phylum", "Subphylum", "Class", "Subclass", "Order", "Suborder", 
    "Family", "Subfamily", "Genus", "Subgenus", "Species", "Exemplar GenBank Accession Number", 
    "Realm", "Subrealm", "Kingdom", "Subkingdom", "Phylum", "Subphylum", 
    "Class", "Subclass", "Order", "Suborder", "Family", "Subfamily", 
    "Genus", "Subgenus", "Species", "Exemplar GenBank Accession Number", 
    "Exemplar virus name", "Virus name abbrevn", "Exemplar isolate designation", 
    "Genome coverage", "Genome composition", "Host/Source", "Change", 
    "Rank", NA_character_, NA, NA, NA, NA_character_, NA_character_, 
    NA_character_, NA_character_, NA_character_)

xlsx_valid_ranks =str_to_lower(c("Realm", "Subrealm", "Kingdom", "Subkingdom", 
    "Phylum", "Subphylum", "Class", "Subclass", "Order", "Suborder", 
    "Family", "Subfamily", "Genus", "Subgenus", "Species"))

#
# merged output colnames
#
xlsx_out_ranks=c("Realm", "Subrealm", "Kingdom", "Subkingdom", 
    "Phylum", "Subphylum", "Class", "Subclass", "Order", "Suborder", 
    "Family", "Subfamily", "Genus", "Subgenus", "Species"
    )
xlsx_out_accession=c("exemplarAccessions")
xlsx_out_other=c( "exemplarName", 
                  "Abbrev", 
                  "exemplarIsolate", 
                  "isComplete", # Genome Coverage
                  "molecule",   # Genome Composition", 
                  "hostSource",
                  "change",
                  "rank",
                  "comments"
)
xlsx_in_colnames = c(  
  paste0("src",c(xlsx_out_ranks,xlsx_out_accession)),
  tolower(xlsx_out_ranks), 
  xlsx_out_accession,
  xlsx_out_other )

xlsx_out_colnames = c(
  "dest_msl_release_num", "filename", "sort","proposal_abbrev","proposal","spreadsheet",
  xlsx_in_colnames,
  "errors"
)

outTemplateDf = setNames(data.frame(matrix(ncol = length(xlsx_out_colnames), nrow = 0)), xlsx_out_colnames)
```
## Build merged proposal

```{r load_and_merge}
mergedDf = setNames(data.frame(matrix(ncol = length(xlsx_out_colnames), nrow = 0)), xlsx_out_colnames)
statusDf = setNames(data.frame(matrix(ncol=6,nrow=0)),c("zip","spreadsheet","nrow","ncol","nchanges","status"))
statusRow =0

# paste two string vectors, preserving double-NAs
# example: pasteNA(x=c(NA,NA,"fred","sam"), y=c(NA,"lucy","george",NA),sep=";") #== c(NA, "lucy","fred;george","sam")
pasteNA = function(x,y,...) {
  ifelse(
    is.na(x) & is.na(y),NA,  # (NA,NA) -> NA
         ifelse(is.na(x),y,  # (NA,y) -> y
                ifelse(is.na(y),x,  # (x,NA) -> x
                       paste(x,y,...) # (x,y) -> paste0(x,y)
                       )
                )
    )
}

# append DF for current proposal to the merged output
mergeProDf = function(mergedDf, pro_df) {
  # concat
  mergedDf = rbind(mergedDf,pro_df[,1:48])
  cat("After ", pro_xls_fname, " mergedDf has ", dim(mergedDf),"\n")
  return(mergedDf)
}
mergeErrorDf = function(mergedDf, spreadsheet_fname, error_str) {
  # make DF for this proposal to hold error
  errorDf = outTemplateDf
  errorDf[1,c("spreadsheet","errors")] = c(spreadsheet_fname, error_str)
  print(errorDf[,c("spreadsheet","errors")])
  mergedDf = mergeProDf(mergedDf, errorDf)
  return(mergedDf)
}

for(pro_zip in rownames(zipDf) ) {
  # pro_zip = rownames(zipDf)[2] # debug
  # pro_zip = rownames(zipDf)[which(pro_zip == rownames(zipDf))+1];which(pro_zip == rownames(zipDf)) # debug "NEXT"
  # pro_zip = "2021.004P.A.v1.Anulavirus_1ns.zip" # incorrect CODE
  # pro_zip = "2021.010P.A.v1.Emaravirus_1ns.zip" # has long-dash in exemplarAccessions
  # pro_zip = '2021.010M.A.v1.Bornaviridae_sprename.zip' # has UTF8 characters
  zip = zipDf[pro_zip,"topZip"] # debug
  # track per-file status
  statusRow = statusRow+1
  statusDf[statusRow,c("zip","spreadsheet","status")] = c(zip,NA,NA)
  
  # extract proposal documents from zip
  
  ### QC that .xlsx filename matches .docx filename matches .zip filename (ignore extra docs and xls's)
  pro_zip_fname=zipDf[pro_zip,"proPath"]#file.path(params$pro_dir,pro_zip)
  pro_toc = unzip(zipfile=pro_zip_fname, list=TRUE)$Name
  pro_toc = grep(pro_toc, pattern="__MACOSX", value = T, invert = T)
  if( params$unzipPro) { unzip(zipfile=pro_zip_fname,files = pro_toc, exdir=zipDf[pro_zip,"xlsxDir"]) }
  pro_xls_fname = zipDf[pro_zip,"xlsx"]
  if( pro_xls_fname %in% pro_toc ) {
    statusDf[statusRow,"spreadsheet"]=pro_xls_fname
  } else{ 
    error_str = "ERROR: missing main .xlsx file"
    cat(error_str," in [",zip,"] ",pro_zip_fname,"\n")
    cat("\t",paste0(pro_toc,collapse="\n\t"),"\n")
    error_str = paste0(error_str,": ",pro_xls_fname, " NOT IN ",pro_zip_fname,"[",paste0(pro_toc,collapse=","),"]")
    # add error entry to output df
    mergedDf=mergeErrorDf(mergedDf,pro_zip_fname,error_str)
    statusDf[statusRow,c("spreadsheet","status")] = c(pro_zip_fname,error_str)
    next
  } 
  
  # load proposal xlsx
  pro_xls_tmp_fname=file.path(zipDf[pro_zip,"xlsxDir"],pro_xls_fname)
  df = data.frame(read_excel(pro_xls_tmp_fname,trim_ws=TRUE,na="Please select",skip=2,range=cell_cols("A:AO"),col_names = FALSE ))
  # human readable column names
  colnames(df)=xlsx_in_colnames
  statusDf[statusRow,c("nrow","ncol")]=c(nrow(df),ncol(df))
  
  # extract code, QC cell(A1)=="Code assigned"
  pro_code_from_xls=df[1,6]
  if(df[1,1] == "Code assigned") {
    cat("QC: ROW1: Ok\n")
    cat("# code_assigned=", pro_code_from_xls, "\n",sep ="")
  } else {
    error_str = paste0("ERROR: ","A1 <> 'Code assigned'; A1='",df[1,1],"'")
    cat(error_str," in [",zip,"] ",pro_zip_fname,"\n")
    # add error entry to output df
    mergedDf=mergeErrorDf(mergedDf,pro_zip_fname,error_str)
    statusDf[statusRow,c("spreadsheet","status")] = c(pro_xls_fname,error_str)
    next
  } 
  
  #
  # QC columns
  #
  row2_errcols = which(as.character(df[2,]) != xlsx_row2[1:ncol(df)])
  if( length(row2_errcols)==0 ) {
    cat("QC: ROW2: OK\n")
  } else {
    error_str = paste0("ERROR: ",
                       "xlsx row 2 modified: ",
                       # add column names
                       paste(colnames(df)[row2_errcols],
                             # mismatched values
                             paste(df[2,row2_errcols],xlsx_row2[row2_errcols],sep="<>")
                             ,sep=":",collapse=";"
                       )
    )
    cat(error_str," in [",zip,"] ",pro_zip_fname,"\n")
    
    # add error to output
    mergedDf=mergeErrorDf(mergedDf,pro_zip_fname,error_str)
    statusDf[statusRow,c("spreadsheet","status")] = c(pro_xls_fname,error_str)
    next
  }
  row3_trimmed = gsub("\\s$","",gsub("\\s+"," ",as.character(df[3,])))
  row3_errcols = which(row3_trimmed != xlsx_row3[1:ncol(df)])
  if( length(row3_errcols)==0  ) {
    cat("QC: ROW3: OK\n")
  } else {
    error_str = paste0("ERROR: ",
                       "xlsx row 3 modified: ",
                       # add column names
                       paste(colnames(df)[row3_errcols],
                             # mismatched values
                             paste(row3_trimmed[row3_errcols],xlsx_row3[row3_errcols],sep="<>")
                             ,sep=":",collapse=";"
                       )
    )
    cat(error_str," in [",zip,"] ",pro_zip_fname,"\n")
    # add error to output
    mergedDf=mergeErrorDf(mergedDf,pro_zip_fname,error_str)
    statusDf[statusRow,c("spreadsheet","status")] = c(pro_xls_fname,error_str)
    break # debug
    #next
  }
  
  #
  # build output df for this proposal
  #
  
  # find last valid row
  changeCol=match("Change", xlsx_row3)
  rankCol  =match("Rank",   xlsx_row3)
  
  changeOk = !(is.na(df[,changeCol]) | df[,changeCol]%in%c("SPECIFY PROPOSED CHANGE", "Change", "Please select"))
  rankOk   = !(is.na(df[,rankCol  ]) | df[,rankCol  ]%in%c("SPECIFY PROPOSED CHANGE", "Rank",   "Please select"))
  okCount  = sum(changeOk==TRUE)
  # QC
  if( sum(changeOk != rankOk,na.rm=T)!=0) {
    error_str = "ERROR: change and rank colomn discord"
    df[changeOk != rankOk,c(changeCol,rankCol)]
    cat(error_str," in [",zip,"] ",pro_zip_fname,"\n")
    
    # add error to output
    mergedDf=mergeErrorDf(mergedDf,pro_zip_fname,error_str)
    statusDf[statusRow,c("spreadsheet","status")] = c(pro_xls_fname,error_str)
    next
  }
  statusDf[statusRow,"nchanges"] = okCount
  
  # add on prefix columns
  pro_xls_number = gsub(pattern="([^.]+[.][^.]+)[.].*",replacement="\\1",x=pro_xls_fname)
  if( pro_xls_number == pro_code_from_xls) {
    cat("QC: CODE OK; xlsx[A1] matches filename\n")
  } else {
    error_str = paste0("ERROR: ","CODE mismatch: xlsx[A1]=",pro_code_from_xls,", filename=",pro_xls_number)
    cat(error_str," in [",zip,"] ",pro_zip_fname,"\n")
    # add error to output
    mergedDf=mergeErrorDf(mergedDf,pro_zip_fname,error_str)
    statusDf[statusRow,c("spreadsheet","status")] = c(pro_xls_fname,error_str)
    next
  }
  
  #### QQQ: QC ID in the spreadsheet vs the one from the filename
  pro_df = cbind(
    dest_msl_release_num = rep(params$dest_msl, okCount), # target MSL
    filename=              rep(params$merged,   okCount), # output file name
    sort=                  rep(NA,              okCount), # arbitrary sort key (row number)
    proposal_abbrev=       rep(pro_xls_number,  okCount), # proposal code
    proposal=              rep(pro_zip,         okCount), # proposal.zip
    spreadsheet=           rep(pro_xls_fname,   okCount), # proposal.xlsx
    df[changeOk,],
    errors=                rep(NA,              okCount)
  )
  
  #
  # QC the controlled vocabularies (CVs) 
  #
  errorList = list()
  errorSummary = ""
  for(cv_name in names(cvList)) { 
    # cv_name =names(cvList)[1]
    errorList[[cv_name]]=   !pro_df[,cv_name] %in% cvList[[cv_name]]
    errorCount=sum(errorList[[cv_name]])
    if( errorCount > 0) { 
      # set error column in offending rows
      pro_df[errorList[[cv_name]],"errors"] = 
        gsub("(\n\r)","*",
             paste(pro_df[errorList[[cv_name]],"errors"],
                   paste(cv_name,
                         pro_df[errorList[[cv_name]],cv_name],sep="="),
                   sep=";")
        )
      # create 1-line summary
      errorSummary=paste0(errorSummary, ";", paste0(cv_name,"=",errorCount))
    }
  }
  cvErrorCt= sum(unlist(lapply(errorList,sum)))
  if( cvErrorCt == 0 ) {
    cat("QC: CVs OK\n")
  } else { 
    statusDf[statusRow,c("spreadsheet","status")] = c(pro_xls_fname,paste0("WARNING: illegal CV terms N=",cvErrorCt))
  }
  # add proposal to master sheet
  mergedDf = mergeProDf(mergedDf, pro_df)
  
  # debug
  #if( pro_xls_number =="2021.006D") {break}
}
```
# additional QC
```{r qc_global}
# 
# scan for inappropriate characters in exemplarAccessions
#
problemsExemplarAccessions = grep(mergedDf$exemplarAccession, pattern = '[–()-]')

```

# QC problems only

```{r qc_errors}
#
# proposal level errors
#
statusErrors = !is.na(statusDf$status)
kable(statusDf[statusErrors,c("spreadsheet","status")],caption="proposal level errors")

#
# taxon level errors
#
mergedErrors = !is.na(mergedDf$errors)
kable(mergedDf[mergedErrors,c("errors")],caption="taxon level errors")


```
# done

```{r final_stats}
#
# add row numbers
#
mergedDf$sort = 1:nrow(mergedDf)

# copy errors into comments, so they hit the DB
mergedDf$comments = pasteNA(mergedDf$comments, mergedDf$errors)

#
# save
#

# native version
write.table(mergedDf,file=file.path(params$merged),sep="\t",row.names = F, na = "", quote = FALSE)
# UTF-16LE (aka Windows Unicode)
write.table(mergedDf,file=file.path(paste0(params$merged,".utf-16le.txt")),fileEncoding = 'UTF-16LE', sep="\t",row.names = F, na = "", quote = FALSE)

print(paste0("Wrote: ", params$merged,": ", paste(dim(mergedDf),collapse=" lines, ")," cols"))
dim(mergedDf)

write.table(statusDf,file=file.path(params$status),sep="\t",row.names = F, na = "")
print(paste0("Wrote: ", params$status,": ", paste(dim(statusDf),collapse=" lines, ")," cols"))
```
